A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-27 paper.tex
Edits to intro, related work, section 2

Reactive systems are ubiquitous in real-world problems such as circuit design,
industrial automation, or device drivers. Automatic synthesis can provide a
\emph{correct by construction} controller for a reactive system from a
specification.  However, the reactive synthesis problem is 2EXPTIME-complete so
naive algorithms are infeasible on even simple systems.

Reactive synthesis is formalised as a game between the \emph{controller} and
its \emph{environment}. In this work we focus on safety games, in which the
controller must prevent the environment from forcing the game into an error
state.  Much of the complexity of reactive synthesis stems from tracking the
set of states in which a player is winning.

There are several techniques that aim to mitigate this complexity by
representing states symbolically.  Historically the most successful technique
has been to use \emph{Binary Decision Diagrams} (BDDs).  BDDs efficiently
represent a relation on a set of game variables but in the worst case the
representation may be exponential. This means that BDDs are not a
one-size-fits-all solution for all reactive synthesis specifications.

Advances in SAT solving technology has prompted research into its applicability
to synthesis as an alternative to BDDs. One approach is to find sets of states
in CNF \cite{bloem2014}. Another approach is to eschew states and focus on
\emph{runs} of the game. Previous work has applied this idea to realizability
of bounded games \cite{narodytska2014} by forming abstract representations of
the game.  In this paper, we extend this idea to unbounded games by
constructing approximate sets of winning states from abstract trees.

@@

A \emph{controller strategy} $\pi^c : (X, L_u) \to L_c$ is a mapping of states
and uncontrollable inputs to controllable labels. A controller strategy is
winning in a bounded game of rank $n$ if all runs $(x_0, u_0, \pi^c(x_0, u_0)),
(x_1, u_1, \pi^c(x_1, u_1)) \dots (x_n, u_n, \pi^c(x_n, u_n))$ are winning.
Bounded \emph{realizability} is the problem of determining the existence of
such a strategy for a bounded game.

@@

An abstract game tree can be checked for the existence of a winning run by a
SAT solver\cite{narodytska2014}.  The tree must be encoded into CNF by making
copies of the transition relation for each game step in tree. The formula must
also check whether the error state has been reached in no branch
(\emph{respectively}: in all branches) for the controller (\emph{resp}:
environment.) The functions that produce these formulas are shown in Algorithm
\ref{alg:treeFormula}.

@@

The \emph{must-invariant} guarantees that the must-losing set $B^M$ is an
underapproximation of the actual losing set $B$: $$B^M \subseteq B$$.

@@

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any
assignment to $X_n$ that satisfies $\sigma(X_T) \land |treeFormula|(T_1)$ also
satisfies $\II$.  Hence, removing $\II$ from $B^m[\textsc{rank}(n)]$ makes
$\sigma(X_T) \land |treeFormula|(T_1)$ unsatisfiable, and hence all
preconditions of the recursive invocation of \textsc{\textoverline{learn}} in
line~(17) are satisfied.

@@

>>>>

Reactive systems are ubiquitous in real-world problems such as circuit design,
industrial automation, or device drivers. Automatic synthesis can provide a
\emph{correct by construction} controller for a reactive system from a
specification.  However, the reactive synthesis problem is 2EXPTIME-complete so
naive algorithms are infeasible on anything but simple systems.

Reactive synthesis is formalised as a game between the \emph{controller} and
its \emph{environment}. In this work we focus on safety games, in which the
controller must prevent the environment from forcing the game into an error
state. In this way the environment is equivalent to an existential search of
the game and the controller is universal. Much of the complexity of reactive
synthesis stems from managing the interactions of the alternating quantifiers
on the state space of the game.

There are several techniques that aim to mitigate this complexity by
representing states symbolically.  Historically the most successful technique
has been to use \emph{Binary Decision Diagrams} (BDDs).  BDDs efficiently
represent a relation on a set of game variables but in the worst case the
representation may be exponential in the number of variables. This means that
BDDs are not a one-size-fits-all solution for all reactive synthesis
specifications.

Advances in SAT solving technology has prompted research into its applicability
to synthesis as an alternative to BDDs. One approach is to find sets of states
in CNF \cite{bloem2014,morgenstern2013}. Another approach is to eschew states
and focus on \emph{runs} of the game. Previous work has applied this idea to
realizability of bounded games \cite{narodytska2014} by forming abstractions of
the game and refining in a counterexample-guided framework. This has been shown
to outperform BDDs on certain classes of game specifications but is only able
to find counterexamples of a certain length to the safety property.  In this
paper, we extend this idea to unbounded games by approximating sets of unsafe
states from abstract games. Careful construction ensures that a fixed point in
these sets guarantees completeness.

Section 2 outlines the original bounded synthesis algorithm. In Section 3 we
describe and prove the correctness of our extension of the algorithm to
unbounded games. In the following sections we discuss optimisations to the
algorithm, evaluate our methodology, and compare our approach to other
synthesis techniques.

@@

A \emph{controller strategy} $\pi^c : (X, L_u) \to L_c$ is a mapping of states
and uncontrollable inputs to controllable labels. A controller strategy is
winning in a bounded game of rank $n$ if all runs $(x_0, u_0, \pi^c(x_0, u_0)),
(x_1, u_1, \pi^c(x_1, u_1)) \dots$ $(x_n, u_n, \pi^c(x_n, u_n))$ are winning.
Bounded \emph{realizability} is the problem of determining the existence of
such a strategy for a bounded game.

@@

An abstract game tree can be checked for the existence of a winning run by a
SAT solver\cite{narodytska2014}.  The tree must be encoded into CNF by making
copies of the transition relation for each game step in tree. The formula must
also check whether the error state has been reached in no branch
(\emph{respectively}: in all branches) for the controller (environment.) The
functions that produce these formulas are shown in Algorithm
\ref{alg:treeFormula}.

@@

The \emph{must-invariant} guarantees that the must-losing set $B^M$ is an
underapproximation of the actual losing set $B$: $$B^M \subseteq B.$$

@@

Furthermore, the interpolant satisfies $F_1 \rightarrow \II$, i.e., any
assignment to $X_n$ that satisfies $\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ also satisfies $\II$.  Hence,
removing $\II$ from $B^m[\textsc{rank}(n)]$ makes $\sigma(X_T) \land
\textsc{\textoverline{treeFormula}}(T_1)$ unsatisfiable, and hence all
preconditions of the recursive invocation of \textsc{\textoverline{learn}} in
line~(17) are satisfied.

@@

The algorithm presented in this paper solves realisability for safety games.
There is a method for extracting strategies from abstract game trees that is
compatible with our method \cite{een2015}. It involves a similar interpolation
approach for discovering states in game tree nodes.
